\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1.1in]{geometry}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\C}{\mathbb{C}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\A}{A_3}
\newcommand{\ord}{\operatorname{ord}}
\newcommand{\gr}{\operatorname{gr}}
\newcommand{\adj}{\operatorname{adj}}

\title{A certified companion to the Dixmier counterexample in $A_3$,\\
and an explicit counterexample to the Poisson conjecture}
\author{Brian Naughton\thanks{In collaboration with Claude Fable 5 (Anthropic), which carried out the
extraction, the proofs and the symbolic certification. Adversarial review
by ChatGPT-5.6 Sol (OpenAI) was a material contribution: it identified a
gap in the original non-surjectivity argument, a sign error, and the
prior-art citations now given prominently. All computations are
reproducible from the repository accompanying this note.}}
\date{21 July 2026}

\begin{document}
\maketitle

\begin{abstract}
On 19 July 2026, Alp\"oge announced an explicit non-injective polynomial map
$F\colon\C^3\to\C^3$ with constant Jacobian determinant $-2$, refuting the
Jacobian conjecture in dimension three~\cite{Alpoge2026}. The same night,
Omniscience Research Agent and Pickhardt~\cite{OP2026} carried $F$ across
the classical inverse-Jacobian pullback to obtain an explicit injective,
non-surjective endomorphism of the third Weyl algebra, refuting the
generalised Dixmier conjecture $DC_n$ for $n\ge3$; priority for that result
is theirs. The present note contributes three things. First, a
\emph{runnable exact-arithmetic certificate}: a short script that verifies,
as polynomial identities over $\Q$, every computational claim on which the
construction rests. Second, the \emph{explicit Poisson witness}: applying
Bavula's same-rank associated-graded implication $PC_n\Rightarrow JC_n$
\cite[Thm.~7]{Bavula2024} to Alp\"oge's map, we display and certify the
resulting injective, non-surjective Poisson endomorphism of the standard
symplectic Poisson algebra $\C[x_1,x_2,x_3,\xi_1,\xi_2,\xi_3]$. The
transfer mechanism is Bavula's; as of 21 July 2026 we are unaware of an
earlier note displaying and verifying this particular witness. Third, a
streamlined self-contained exposition of the transfer. If Zheglov's
announced proof of $DC_1$~\cite{Zheglov2024} is confirmed, $DC_2$ is the
only undecided rank.
\end{abstract}

\section{Introduction}

Let $A_n = A_n(\C)$ denote the $n$-th Weyl algebra: the unital associative
$\C$-algebra with generators $x_1,\dots,x_n,\partial_1,\dots,\partial_n$ and
relations
\begin{equation}\label{eq:ccr}
[x_i,x_j]=0,\qquad [\partial_i,\partial_j]=0,\qquad
[\partial_i,x_j]=\delta_{ij}\qquad(1\le i,j\le n).
\end{equation}
Dixmier~\cite{Dixmier1968} asked in 1968 whether every algebra endomorphism
of $A_1$ is an automorphism; the generalised \emph{Dixmier conjecture}
$DC_n$ asserts this for $A_n$. It is classical that $DC_n$ implies the
Jacobian conjecture $JC_n$ in the same dimension (see van den
Essen~\cite{vdE2000}; the construction of the pullback appears in
Bavula~\cite{Bavula2024}, and the order-preservation argument is recorded by
Belov-Kanel and Kontsevich~\cite[p.~2]{BKK2007}), while the much deeper
converse passes to doubled dimension: $JC_{2n}\Rightarrow
DC_n$~\cite{Tsuchimoto2005,BKK2007}; see also Adjamagbo--van den
Essen~\cite{AvdE2007}, who add the Poisson conjecture to the equivalence
web.

On 19 July 2026, Alp\"oge, working with the language model Claude Fable~5,
announced an explicit counterexample to $JC_3$~\cite{Alpoge2026}: the map
$F=(F_1,F_2,F_3)\colon\C^3\to\C^3$,
\begin{equation}\label{eq:F}
\begin{aligned}
F_1 &= (1+xy)^3 z + y^2(1+xy)(4+3xy),\\
F_2 &= y + 3x(1+xy)^2 z + 3xy^2(4+3xy),\\
F_3 &= 2x - 3x^2y - x^3 z,
\end{aligned}
\end{equation}
satisfies $\det JF \equiv -2$ yet identifies three distinct points:
\begin{equation}\label{eq:collision}
F\bigl(0,\,0,\,-\tfrac14\bigr)
= F\bigl(1,\,-\tfrac32,\,\tfrac{13}{2}\bigr)
= F\bigl(-1,\,\tfrac32,\,\tfrac{13}{2}\bigr)
= \bigl(-\tfrac14,\,0,\,0\bigr).
\end{equation}
That same night, Omniscience Research Agent and Pickhardt~\cite{OP2026}
published the resulting explicit counterexample to $DC_3$, with two
complete proofs of non-surjectivity (via centralisers and via the order
filtration), a Gr\"obner-basis verification that the fibre over
$(-\tfrac14,0,0)$ consists of exactly the three points
\eqref{eq:collision}, and the rank monotonicity $DC_n\Rightarrow DC_m$ for
$m\le n$. \textbf{Priority for the Dixmier counterexample belongs
to~\cite{OP2026}}; we learned of their preprint after completing an
independent derivation of the same result, and this note has been reframed
accordingly.

What this note adds:
\begin{enumerate}
\item \textbf{A certificate} (Section~\ref{sec:certified}). Reference
\cite{OP2026} observes that everything can ``be confirmed by exact
arithmetic rather than taken on faith'' but ships no artefact for doing so.
The repository accompanying this note contains a short script certifying,
as exact polynomial identities over $\Q$, every computational ingredient:
the determinant, the collision, the Weyl relations for the pullback, and
the Poisson relations for its graded shadow.
\item \textbf{The Poisson instance} (Section~\ref{sec:poisson}). The
associated-graded map of the Weyl pullback is an explicit injective,
non-surjective endomorphism of the polynomial Poisson algebra
$\C[x,\xi]$ with its standard symplectic bracket, refuting the Poisson
conjecture (as formulated in~\cite{AvdE2007}) in rank $\ge3$ with an
explicit witness. The same-rank transfer mechanism --- extend the Keller
map to the Weyl algebra, pass to the associated graded, observe a Poisson
endomorphism --- is prior art: it is exactly Bavula's proof of
$PC_n\Rightarrow JC_n$~\cite[Thm.~7]{Bavula2024}. What we contribute is
the written-out, certified instance at Alp\"oge's map, a consequence not
displayed in \cite{OP2026} or, to our knowledge, in the other early
consequence notes (e.g.~\cite{Zhang2026}, which derives Mathieu-type
consequences).
\item \textbf{A compact exposition} (Sections~\ref{sec:weyl}--\ref{sec:nonsurj}),
kept because the certificate refers to its numbered identities.
\end{enumerate}

\section{The map, and what is certified}\label{sec:certified}

Write $x=(x_1,x_2,x_3)=(x,y,z)$ and $J = JF = (\partial F_i/\partial
x_k)_{ik}$. Since $\det J\equiv-2$ (an identity of polynomials; certified,
and provable by hand via the elegant substitution of \cite[Lem.~1]{OP2026}),
the matrix
\[
G \;:=\; (J^{\mathsf T})^{-1} \;=\; -\tfrac12\,\adj(J^{\mathsf T})
\]
has entries in $\Q[x,y,z]$, with $\det G = 1/\det J^{\mathsf T} = -\tfrac12$.
The accompanying script \texttt{verify\_endomorphism.py} certifies, by exact
symbolic computation over $\Q$ (every check is a polynomial in
$\Q[x,y,z]$ or $\Q[x,\xi]$ tested for literal vanishing):
\begin{itemize}
\item[(C0)] $\det J \equiv -2$; the three points in \eqref{eq:collision}
are pairwise distinct and share the stated image; $\det G \equiv -\tfrac12$;
\item[(R1)] $G\,J^{\mathsf T} = I$ \quad (nine identities);
\item[(R2)] $\displaystyle\sum_{k}\Bigl(G_{ik}\,\partial_k G_{jl} -
G_{jk}\,\partial_k G_{il}\Bigr) = 0$ for all $i,j,l$ \quad (the flatness
identities);
\item[(P1--P3)] the Poisson relations of Section~\ref{sec:poisson}, checked
directly against the definition of the bracket.
\end{itemize}
The script prints its Python and SymPy versions, exits successfully only if
every identity holds, and writes the nine entries of $G$ to a
machine-readable file \texttt{G.json}, so that the endomorphism below is
explicit without any re-derivation by the reader.

\section{The Weyl pullback}\label{sec:weyl}

Define, following the classical construction (Bavula~\cite{Bavula2024};
\cite[\S3]{OP2026}):
\[
\varphi(x_i) = F_i,\qquad
\varphi(\partial_j) = D_j := \sum_{k=1}^{3} G_{jk}\,\partial_k .
\]

\begin{proposition}\label{prop:welldef}
$\varphi$ extends uniquely to a unital algebra endomorphism of $\A$.
\end{proposition}

\begin{proof}
By the presentation \eqref{eq:ccr} it suffices to check the relations on
the images. The $F_i$ commute. Since $F_i$ is a multiplication operator,
$[D_j,F_i]=D_j(F_i)=\sum_k G_{jk}\,\partial F_i/\partial x_k=(G\,J^{\mathsf
T})_{ji}=\delta_{ji}$ by (R1). For (R2) there is both the certificate and
the conceptual argument of \cite[Prop.~3]{OP2026}: $[D_i,D_j]$ is a
derivation $\sum_l c_l\partial_l$ of $\C[x,y,z]$, and applying it to $F_m$
gives $[D_i,D_j](F_m)=D_i(\delta_{jm})-D_j(\delta_{im})=0$, i.e.\
$J\,(c_1,c_2,c_3)^{\mathsf T}=0$; invertibility of $J$ over $\C[x,y,z]$
forces every $c_l=0$.
\end{proof}

\section{Injectivity and failure of surjectivity}\label{sec:nonsurj}

Every element of $\A$ has a unique normal form $P=\sum_\alpha
a_\alpha(x)\,\partial^\alpha$; for $P\neq0$ set $\ord(P) := \max\{|\alpha| :
a_\alpha\neq0\}$ and $\Sigma_m=\{P:\ord(P)\le m\}\cup\{0\}$. The associated
graded algebra is the commutative polynomial ring
$\gr\A=\C[x_1,x_2,x_3][\xi_1,\xi_2,\xi_3]$~\cite[Ch.~2--3]{Coutinho1995}.
We identify $\xi_j$ with $-\sigma(\partial_j)$, where $\sigma$ denotes the
principal symbol; with this identification the graded bracket determined
by $\{\sigma(P),\sigma(Q)\}=\sigma([P,Q])$ is the standard symplectic
bracket of Section~\ref{sec:poisson}, with $\{x_i,\xi_j\}=+\delta_{ij}$,
and no sign disclaimers are needed below.

\begin{lemma}\label{lem:algind}
$F_1,F_2,F_3$ are algebraically independent over $\C$; equivalently
$a\mapsto a\circ F$ is injective on $\C[x,y,z]$.
\end{lemma}

\begin{proof}
Standard in characteristic zero from $\det J\ne 0$. A nonzero constant $Q$
cannot satisfy $Q(F)=0$, so a nonzero relation $Q$ of minimal total degree
has positive degree and some $\partial Q/\partial y_i\ne0$. Applying
$\partial/\partial x_k$ to $Q(F)=0$ gives $\sum_i (\partial Q/\partial
y_i)(F)\, J_{ik}=0$; invertibility of $J$ over $\C(x,y,z)$ forces
$(\partial Q/\partial y_i)(F)=0$ for all $i$, contradicting minimality.
\end{proof}

\begin{lemma}\label{lem:filtration}
$\varphi(\Sigma_m)\subseteq\Sigma_m$, the induced graded map is
\[
(\gr\varphi)(P)(x,\xi) \;=\; P\bigl(F(x),\, G(x)\,\xi\bigr),
\]
$\gr\varphi$ is injective, and consequently
$\ord(\varphi(P))=\ord(P)$ for every nonzero $P\in\A$.
\end{lemma}

\begin{proof}
$\varphi(x_i)\in\Sigma_0$ and $\varphi(\partial_j)\in\Sigma_1$, so
$\varphi$ is filtered and $\gr\varphi$ is the displayed map on
$\gr\A=\C[x][\xi]$. Suppose $P(F,G\xi)=0$. Over $K=\C(x,y,z)$ the matrix
$G$ is invertible ($\det G=-\tfrac12$), so substituting
$\xi\mapsto G^{-1}\xi$ gives $P(F(x),\xi)=0$ in $K[\xi]$; reading off
$\xi$-coefficients and applying Lemma~\ref{lem:algind} gives $P=0$. Strict
order preservation follows: a nonzero principal symbol has nonzero image.
\end{proof}

\begin{theorem}[{\cite[Cor.~7]{OP2026}}; independent derivation]\label{thm:weyl}
$\varphi$ is injective and not surjective. Hence $DC_3$ is false, and by
padding $F$ with identity coordinates, $DC_n$ is false for every $n\ge 3$.
\end{theorem}

\begin{proof}
$\A$ is simple~\cite{Dixmier1968}, so the kernel of the unital endomorphism
$\varphi$ is zero. In fact $x_1\notin\operatorname{im}\varphi$: if
$\varphi(P)=x_1$ then Lemma~\ref{lem:filtration} gives $\ord(P)=0$, so
$P=p(x)$ is a polynomial with $p(F)=x_1$. Evaluating at the first two
points of \eqref{eq:collision}, whose images under $F$ coincide, gives
$p(-\tfrac14,0,0)=0$ and $p(-\tfrac14,0,0)=1$, a contradiction.
\end{proof}

\section{The explicit Poisson counterexample}\label{sec:poisson}

Equip $\mathcal P_3 := \C[x_1,x_2,x_3,\xi_1,\xi_2,\xi_3]$ with the standard
symplectic Poisson bracket, determined by
\[
\{x_i,x_j\}=0,\qquad \{\xi_i,\xi_j\}=0,\qquad \{x_i,\xi_j\}=\delta_{ij},
\qquad
\{f,g\}=\sum_k\Bigl(\frac{\partial f}{\partial x_k}\frac{\partial
g}{\partial \xi_k}-\frac{\partial f}{\partial \xi_k}\frac{\partial
g}{\partial x_k}\Bigr).
\]
The \emph{Poisson conjecture} in rank $n$ (as formulated
in~\cite{AvdE2007}) asserts that every Poisson endomorphism of
$\mathcal P_n$ --- a $\C$-algebra endomorphism preserving the bracket ---
is an automorphism.

\begin{theorem}\label{thm:poisson}
The assignment
\[
\psi(x_i) = F_i,\qquad
\psi(\xi_j) = \eta_j := \sum_{k=1}^3 G_{jk}\,\xi_k
\]
extends uniquely to a Poisson endomorphism $\psi$ of $\mathcal P_3$
(equivalently, $\psi=\gr\varphi$), and $\psi$ is injective but not
surjective. Hence the Poisson conjecture is false in rank $3$, and, by
padding, in every rank $n\ge 3$.
\end{theorem}

\begin{proof}
\emph{Well defined.} $\psi$ is a polynomial-algebra endomorphism on the
$6$ generators; it is Poisson iff the brackets of the images match the
brackets of the generators. $\{F_i,F_j\}=0$ since the $F_i$ do not involve
$\xi$ (P1). $\{F_i,\eta_j\}=\sum_k (\partial F_i/\partial x_k)
G_{jk}=(G J^{\mathsf T})_{ji}=\delta_{ij}$ by (R1) (P2). Finally
$\{\eta_i,\eta_j\}=\sum_l\bigl(\sum_k G_{jk}\,\partial_k G_{il}-
G_{ik}\,\partial_k G_{jl}\bigr)\xi_l$, whose $l$-th coefficient is the
\emph{negative} of the flatness expression (R2), hence vanishes (P3). All
three families are also certified directly against the definition of the
bracket by the script.

\emph{Injective.} $\psi$ is exactly the map $P\mapsto P(F,G\xi)$ of
Lemma~\ref{lem:filtration}, shown there to be injective.

\emph{Not surjective.} $\psi$ is graded for the $\xi$-degree: $F_i$ has
$\xi$-degree $0$ and $\eta_j$ has $\xi$-degree $1$, so $\psi$ maps
$\xi$-degree-$m$ homogeneous elements to $\xi$-degree-$m$ homogeneous
elements. Suppose $\psi(P)=x_1$ and write $P=\sum_m P_m$ with $P_m$
homogeneous of $\xi$-degree $m$. Since $x_1$ has $\xi$-degree $0$, we get
$\psi(P_m)=0$ for every $m>0$; injectivity gives $P_m=0$, so $P=p(x)$ is a
polynomial with $p(F)=x_1$. Evaluating at the first two points of
\eqref{eq:collision}, whose images under $F$ coincide, gives
$p(-\tfrac14,0,0)=0$ and $p(-\tfrac14,0,0)=1$, a contradiction. Hence
$x_1\notin\operatorname{im}\psi$.

For $n\ge4$, pad $F$ with identity coordinates as in
Theorem~\ref{thm:weyl}; the same argument applies verbatim.
\end{proof}

\begin{remark}
The transfer mechanism in Theorem~\ref{thm:poisson} is not new: extending
a Keller map to the Weyl algebra, passing to the associated graded, and
observing a Poisson endomorphism whose invertibility would force
invertibility of the original map is precisely Bavula's same-rank proof of
$PC_n\Rightarrow JC_n$~\cite[Thm.~7]{Bavula2024}; no stable equivalence
and no rank doubling enters. Our contribution is the explicit certified
instance at the map \eqref{eq:F}. With the identification of
Section~\ref{sec:nonsurj}, $\psi=\gr\varphi$ literally: the counterexample
to the quantum statement and the counterexample to its classical limit are
associated graded to one another.
\end{remark}

\begin{remark}[Status of the remaining ranks]
Rank monotonicity ($DC_n\Rightarrow DC_m$ for $m\le n$,
\cite[Prop.~9]{OP2026}) means the refutation propagates upward only.
Zheglov has posted a preprint claiming a proof of
$DC_1$~\cite{Zheglov2024}; we do not assess its status. If that claim is
confirmed, $DC_2$ is the only undecided rank of the Dixmier family, as
$JC_2$ is of the Jacobian family.
\end{remark}

\subsection*{Code availability}
\texttt{verify\_endomorphism.py} and \texttt{G.json} accompany this note.
A kernel-checked (Lean~4) rendering of the certified identities is
mechanical and planned as a follow-up; we make no claim for it here.

\subsection*{Acknowledgements}
The counterexample $F$ is due to Levent Alp\"oge with Claude
Fable~5~\cite{Alpoge2026}. The Dixmier counterexample was first published
by Omniscience Research Agent and Jeff Pickhardt~\cite{OP2026}, whose
careful preprint we warmly acknowledge; the present note was derived
independently on 21 July and reframed as a companion upon learning of
theirs. We thank ChatGPT-5.6 Sol (OpenAI) for two rounds of adversarial review
that materially improved this note: the first caught a sign error in
$\det G$ and the missing prior-art citation; the second identified a
genuine gap in the original non-surjectivity argument (repaired in
Theorems~\ref{thm:weyl} and~\ref{thm:poisson} with a stronger direct
proof) and the correct attribution of the same-rank transfer to
Bavula~\cite[Thm.~7]{Bavula2024}. Errors that remain are ours.

\begin{thebibliography}{9}

\bibitem{Dixmier1968}
J.~Dixmier, \emph{Sur les alg\`ebres de Weyl}, Bull. Soc. Math. France
\textbf{96} (1968), 209--242.

\bibitem{Alpoge2026}
L.~Alp\"oge, X post displaying the map \eqref{eq:F} and the collision
\eqref{eq:collision}, 19 July 2026,
\texttt{x.com/\_\_alpoge\_\_/status/2079028340955197566}.

\bibitem{OP2026}
Omniscience Research Agent and J.~Pickhardt, \emph{An explicit
counterexample to the Dixmier conjecture in $A_3$}, 19 July 2026,
\texttt{omniscienceproject.com/papers/an-explicit-counterexample-to-the-\\
dixmier-conjecture-in-a-3-jfLENtXF}.

\bibitem{Bavula2024}
V.~V.~Bavula, \emph{Holonomic modules and 1-generation in the Jacobian
Conjecture}, C. R. Math. \textbf{362} (2024), 731--738; arXiv:2112.03177.

\bibitem{vdE2000}
A.~van den Essen, \emph{Polynomial Automorphisms and the Jacobian
Conjecture}, Progress in Mathematics 190, Birkh\"auser, 2000.

\bibitem{Tsuchimoto2005}
Y.~Tsuchimoto, \emph{Endomorphisms of Weyl algebra and $p$-curvatures},
Osaka J. Math. \textbf{42} (2005), 435--452.

\bibitem{BKK2007}
A.~Belov-Kanel and M.~Kontsevich, \emph{The Jacobian conjecture is stably
equivalent to the Dixmier conjecture}, Mosc. Math. J. \textbf{7} (2007),
209--218.

\bibitem{AvdE2007}
P.~K.~Adjamagbo and A.~van den Essen, \emph{A proof of the equivalence of
the Dixmier, Jacobian and Poisson conjectures}, Acta Math. Vietnam.
\textbf{32} (2007), 205--214.

\bibitem{Zheglov2024}
A.~Zheglov, \emph{The conjecture of Dixmier for the first Weyl algebra is
true}, preprint, arXiv:2410.06959 (v5, 19 January 2026).

\bibitem{Zhang2026}
Z.~Zhang, \emph{Direct consequences of the three-dimensional counterexample
to the Jacobian conjecture}, note, 20 July 2026.

\bibitem{Coutinho1995}
S.~C.~Coutinho, \emph{A Primer of Algebraic $D$-modules}, LMS Student Texts
33, Cambridge University Press, 1995.

\end{thebibliography}

\end{document}
